Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Affective Values

### 2021-01-14 Dissonance_Santiago_v3/Dissonance_Santiago_v3.tex
v3 Revisions after comments from anonymous reviewers, Journal of Transport and Health

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Affective Values},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

A definition of subjective wellbeing is as ``{[}g{]}ood mental states,
including all of the various evaluations, positive and negative, that
people make of their lives, and the affective reactions of people to
their experiences'' (OECD, 2013, p. 10). As memorably put by Steg
(2005): is the use of a car a must, or is it a lust? A close alignment,
or consonance, between affective reactions and the mode of
transportation used to travel can result in greater subjective wellbeing
and increase the probability of choosing that mode (i.e., by satisfying
the \emph{lust}); on the other hand, dissonance, namely the lack of
concordance between the mode used and the affective reaction to that
mode, would be detrimental to subjective wellbeing, especially to those
for whom the use of the mode is a \emph{must} (De Vos, 2019; Mokhtarian
and Pendyala, 2018). With these considerations in mind, the present
research aims to investigate \emph{who} experiences affective dissonance
with respect to their primary mode of travel. We focus our investigation
on reactions from the perspective of feelings of freedom, enjoyment,
happiness, poverty, luxury, and status - concepts termed \emph{affective
values} hereafter. Furthermore, it is possible, even if use of a
particular mode is a \emph{must}, that travelers may still \emph{lust}
for something else - in other words, the grass may look greener from the
window of a car. For this reason, we also aim to investigate which mode
or modes are most commonly identified as ideal from the perspective of
the aforementioned affective values.

Our research is based on data collected from a sample of travelers in
the city of Santiago in Chile. The analysis sheds light on the affective
reactions of people towards their primary mode of travel, and their
perceptions towards `ideal modes'. This reveals important differences in
situations when the ideal mode of a traveler is not the mode they
actually use - or in other words, the affective dissonance between
actual and ideal modes of travel. More concretely, the results indicate
that users of public transportation experience the most dissonance,
followed by automobile users, and finally active travelers, who tend to
experience the least affective dissonance. For those travelers who
experience dissonance, active travel is the mode most commonly
associated with freedom, enjoyment, and happiness, whereas public
transportation is most commonly associated with poverty, and the
automobile is most commonly associated with luxury and status. We also
find that there are significant variations in dissonance by age,
education, income, and typical commute time. The reactions of travelers
towards various transport modes are critical factors that policy-makers
need to consider to promote and increase the use of public transport and
active modes (Bornioli et al., 2019; De Vos et al., 2019; De Vos and
Witlox, 2017; Garling et al., 2019; Redman et al., 2013).

@@

One of the primary ways to explore issues of subjective wellbeing in
transportation has been the satisfaction that travelers feel towards
their everyday mobility experience (e.g., Cecilia Jakobsson Bergstad et
al., 2011). As a consequence, there is a wealth of research on
satisfaction with the use of different modes of transportation. For
example, numerous studies report that car users often have a higher
level of satisfaction compared to other transport modes (C. J. Bergstad
et al., 2011; Eriksson et al., 2013; Redmond and Mokhtarian, 2001),
although others report highest commute satisfaction for bicycle and
train commuters (e.g., Whalen et al., 2013; Handy and Thigpen, 2019). In
a similar way, there are multiple reports that active travel also tends
to yield higher levels of satisfaction (Gatersleben and Uzzell, 2007;
Handy and Thigpen, 2019; Paez and Whalen, 2010; Smith, 2017; St-Louis et
al., 2014; Whalen et al., 2013), In contrast, public transport users
often assess their experience more negatively (Abenoza et al., 2017; De
Vos et al., 2016; Gatersleben and Uzzell, 2007; Handy and Thigpen, 2019;
Paez and Whalen, 2010). Multi-modal trips also influence satisfaction
levels; for instance, individuals tend to report higher levels of
satisfaction with the mode they use -- perhaps as a form of \emph{post
hoc} validation of their choices or circumstance (Susilo and Cats,
2014).

@@

The literature on SWB has demonstrated a relationship between people's
perceptions and satisfaction with their daily travel (e.g., Smith, 2017;
Mokhtarian and Pendyala, 2018; St-Louis et al., 2014). Scholars have
shown that accessibility is among the most studied factors that
influences subjective wellbeing (Delbosc, 2012), and activities have a
direct impact on travel satisfaction (Cecilia Jakobsson Bergstad et al.,
2011). Delbosc (2012, p. 28), for instance, has summarized the most
significant influences on psychological wellbeing: poverty and
employment, meaningful relationships and health. However, understanding
the components that affect these perceptions implies the differentiation
between affective (also named as symbolic-affective) and instrumental
values (C. J. Bergstad et al., 2011). Steg et al.~(2011) have compared
symbolic-affective and instrumental-reasoned motives for car-use, and
similar findings are reported in other studies (e.g., Gatersleben and
Uzzell, 2007; Lois and Lopez-Saez, 2009). Previous research also
demonstrates how socio-demographic factors affect the levels of SWB,
including income (Clark and Oswald, 1996; Ferrer-i-Carbonell, 2005),
education and unemployment (Argyle et al., 1999), age (Diener and
Eunkook Suh, 1997), and gender (Tesch-Römer et al., 2008). However,
further research beyond the determinants of trip satisfaction is needed
to understand how these socio-demographic variables connect with the
affective reactions to various modes of travel (St-Louis et al., 2014).

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/plot-descriptive-statistics-1.pdf}
\caption{\label{fig:descriptive-statistics}Descriptive statistics of the
sample}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/figure-primary-mode-travel-1.pdf}
\caption{\label{fig:primary-mode-travel}Frequency of primary mode used
for regular commute; top panel: all modes, bottom panel: aggregated
modes}
\end{figure}

@@

As seen in the table, there are five individual attributes that we
analyze. Three of these are socio-economic and demographic, namely age,
level of education, and income. The other two are transportation
related, namely primary mode of travel and commute time. The frequency
tables are tested in every case using \(\chi^2\) tests of independence
(\(p\)-values are reported in the table; lower \(p\)-values mean that
the null hypothesis of independence can be rejected with greater
confidence). It is interesting to note that the only category for which
all affective values are significant at better than 5\% level of
confidence is mode of travel.

@@

Turning now to typical commute time, four affective values show
significant differences at better than 10\% confidence: Freedom,
Happiness, Luxury, and Status. Perhaps not surprisingly, dissonance is
more frequent among people whose typical commutes are longer. This is in
line with previous findings: both St-Louis et al.~(2014) and Smith
(2017) report that commute satisfaction tends to decline with longer
commutes, whereas Handy and Thigpen (2019) found that commute distance
was a negative covariate of commute satisfaction.

@@

> 54 & 74.00 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 72.92 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 68.75 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 76.19 & \multirow{-3}{*}{\centering\arraybackslash 0.4095} & 78.57 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 66.67 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Education}}\\
\hspace{1em}K-12 & 85.98 &  & 92.16 &  & 89.22 &  & 79.27 &  & 95.10 &  & 92.39 & \\

@@

Graduate & 78.57 & \multirow{-3}{*}{\centering\arraybackslash 0.9063} & 85.19 & \multirow{-3}{*}{\centering\arraybackslash 0.7005} & 84.62 & \multirow{-3}{*}{\centering\arraybackslash 0.9694} & 86.36 & \multirow{-3}{*}{\centering\arraybackslash 0.3608} & 76.00 & \multirow{-3}{*}{\centering\arraybackslash 0.0058} & 79.17 & \multirow{-3}{*}{\centering\arraybackslash 0.013}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Income}}\\
\hspace{1em}Low & 86.51 &  & 82.11 &  & 88.62 &  & 80.19 &  & 88.71 &  & 88.70 & \\

@@

High & 79.19 & \multirow{-3}{*}{\centering\arraybackslash 0.5755} & 90.97 & \multirow{-3}{*}{\centering\arraybackslash 0.2264} & 85.82 & \multirow{-3}{*}{\centering\arraybackslash 0.9698} & 93.69 & \multirow{-3}{*}{\centering\arraybackslash 0.0642} & 73.68 & \multirow{-3}{*}{\centering\arraybackslash 0.0204} & 70.31 & \multirow{-3}{*}{\centering\arraybackslash 0.0137}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Mode}}\\
\hspace{1em}Car & 58.93 &  & 78.90 &  & 70.91 &  & 97.96 &  & 41.00 &  & 30.69 & \\

@@

Other & 91.30 & \multirow{-4}{*}{\centering\arraybackslash $< 0.001$} & 86.96 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 91.30 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 93.33 & \multirow{-4}{*}{\centering\arraybackslash 0.0045} & 72.73 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 90.00 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Commute Time}}\\
\hspace{1em}< 20 min & 65.93 &  & 82.95 &  & 77.27 &  & 86.11 &  & 73.49 &  & 67.09 & \\

@@

> 60 min & 95.93 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 92.17 & \multirow{-4}{*}{\centering\arraybackslash 0.6115} & 93.28 & \multirow{-4}{*}{\centering\arraybackslash 0.0394} & 83.02 & \multirow{-4}{*}{\centering\arraybackslash 0.9225} & 91.15 & \multirow{-4}{*}{\centering\arraybackslash 0.0925} & 90.09 & \multirow{-4}{*}{\centering\arraybackslash 0.0158}\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/figure-bar-plots-by-attribute-1.pdf}
\caption{\label{fig:bar-plots-by-attribute}Plots for affective values;
in the y-axis are the number of cases by primary mode of transportation,
and in the x-axis are the modes selected for each affective value}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/figure-bar-plots-by-attribute-and-age-1.pdf}
\caption{\label{fig:bar-plots-by-age}Plots for affective values by age;
in the y-axis is the proportion in the interval {[}0, 1{]}, and the
x-axis is the mode selected for each value (p-values are for
Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/figure-bar-plots-by-attribute-and-education-1.pdf}
\caption{\label{fig:bar-plots-by-education}Plots for affective values by
level of education; in the y-axis is the proportion in the interval
{[}0, 1{]}, and the x-axis is the mode selected for each value (p-values
are for Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/figure-bar-plots-by-attribute-and-income-1.pdf}
\caption{\label{fig:bar-plots-by-income}Plots for affective values by
level of income; in the y-axis is the proportion in the interval {[}0,
1{]}, and the x-axis is the mode selected for each value (p-values are
for Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

Among car users, Freedom is associated with the car in the case of every
commute but the longest (\textgreater60 min), in which case active
travel is more frequently identified as the mode that evokes feelings of
Freedom. The associations of active travel with Enjoyment and Happiness
are consistently high among car users, although car becomes more
frequently associated with this value for longer commutes. As we have
seen before, active travelers seldom experience dissonance with respect
to Freedom, Enjoyment, and Happiness, and if anything the strength of
this association increases with longer commutes. Public transportation
users again are split in their preference towards active travel and car
as the modes more likely to evoke feelings of Freedom, Enjoyment, and
Happiness, and there are only relatively minor differences in these
preferences by length of commute. The last affective value that displays
significant differences in dissonance by length of typical commute is
Status. The car tends to dominate as the mode more frequently associated
with this value among car and public transportation users. Active
travelers, on the other hand, associate status more strongly with short
or very long commutes: only when commutes are shorter than 20 minutes or
longer than 60 minutes do active travelers associate feelings of Status
with the car.

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/figure-bar-plots-by-attribute-and-travel-time-1.pdf}
\caption{\label{fig:bar-plots-by-travel-time}Plots for affective values
by commute time; in the y-axis is the proportion in the interval {[}0,
1{]}, and the x-axis is the mode selected for each value (p-values are
for Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

The premise of this study is that a key component of subjective
wellbeing is the affective reaction to experience (see the definition of
wellbeing offered by the OECD; -@ OECD, 2013). In this paper we
investigated mode dissonance from the perspective of six affective
values. The research presented here was based on a sample of travelers
in Santiago, the capital of Chile. Participants in this research were
asked about their typical mode of travel, and then about the mode or
modes that they associate with each of the six affective values.
Analysis using hypothesis testing (tests of independence) and
visualization techniques (bar charts with faceting) uncovered
interesting patterns. Some of our findings are well aligned with
previous research; for example, active travelers experience less
dissonance than car users, and users of public transportation experience
the most dissonance of all. However, by considering affective values
separately instead of aggregating them into a single indicator of
subjective wellbeing, we manage to preserve greater granularity with
respect to various responses than most studies. This is important
because hedonic/eudaimonic values are more frequently related to active
travel (Freedom, Enjoyment, Happiness), whereas Poverty is more
frequently related to public transportation and active travel. Luxury
and Status, on the other hand, are more frequently associated to car.

@@

\leavevmode\hypertarget{ref-Bergstad2011subjective}{}%
Bergstad, C.J., Gamble, A., Gärling, T., Hagman, O., Polk, M., Ettema,
D., Friman, M., Olsson, L.E., 2011. Subjective well-being related to
satisfaction with daily travel. Transportation 38, 1--15.

\leavevmode\hypertarget{ref-Bergstad2011affective}{}%
Bergstad, C.J., Gamble, A., Hagman, O., Polk, M., Garling, T., Olsson,
L.E., 2011. Affective-symbolic and instrumental-independence
psychological motives mediating effects of socio-demographic variables
on daily car use. Journal of Transport Geography 19, 33--38.
doi:\href{https://doi.org/10.1016/j.jtrangeo.2009.11.006}{10.1016/j.jtrangeo.2009.11.006}

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Affective Values},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

A definition of subjective wellbeing is as ``{[}g{]}ood mental states,
including all of the various evaluations, positive and negative, that
people make of their lives, and the affective reactions of people to
their experiences'' (OECD, 2013, p. 10). As memorably put by Steg
(2005): is the use of a car a must, or is it a lust? A close alignment,
or consonance, between affective reactions and the mode of
transportation used to travel can result in greater subjective wellbeing
and increase the probability of choosing that mode (i.e., by satisfying
the \emph{lust}); on the other hand, dissonance, namely the lack of
concordance between the mode used and the affective reaction to that
mode, would be detrimental to subjective wellbeing, especially to those
for whom the use of the mode is a \emph{must} (De Vos, 2019; Mokhtarian
and Pendyala, 2018). With these considerations in mind, the present
research aims to investigate \emph{who} experiences affective dissonance
with respect to their primary mode of travel. We focus our investigation
on reactions from the perspective of feelings of Freedom, Enjoyment,
Happiness, Poverty, Luxury, and Status - concepts termed \emph{affective
values} hereafter. Furthermore, it is possible, even if use of a
particular mode is a \emph{must}, that travelers may still \emph{lust}
for something else - in other words, the grass may look greener from the
window of a car. For this reason, we also aim to investigate which mode
or modes are most commonly identified as ideal from the perspective of
the aforementioned affective values.

Our research is based on data collected from a sample of travelers in
the city of Santiago in Chile. The analysis sheds light on the affective
reactions of people towards their primary mode of travel, and their
perceptions towards `ideal modes'. This reveals important differences in
situations when the ideal mode of a traveler is not the mode they
actually use - or in other words, the affective dissonance between
actual and ideal modes of travel. More concretely, the results indicate
that users of public transportation experience the most dissonance,
followed by automobile users, and finally active travelers, who tend to
experience the least affective dissonance. For those travelers who
experience dissonance, active travel is the mode most commonly
associated with Freedom, Enjoyment, and Happiness, whereas public
transportation is most commonly associated with Poverty, and the
automobile is most commonly associated with Luxury and Status. We also
find that there are significant variations in dissonance by age,
education, income, and typical commute time. The reactions of travelers
towards various transport modes are critical factors that policy-makers
need to consider to promote and increase the use of public transport and
active modes (Bornioli et al., 2019; De Vos et al., 2019; De Vos and
Witlox, 2017; Garling et al., 2019; Redman et al., 2013).

@@

One of the primary ways to explore issues of subjective wellbeing in
transportation has been the satisfaction that travelers feel towards
their everyday mobility experience (e.g., Jakobsson Bergstad, Gamble, T.
Gärling, et al., 2011). As a consequence, there is a wealth of research
on satisfaction with the use of different modes of transportation. For
example, numerous studies report that car users often have a higher
level of satisfaction compared to other transport modes (Eriksson et
al., 2013; Jakobsson Bergstad, Gamble, Hagman, et al., 2011; Redmond and
Mokhtarian, 2001), although others report highest commute satisfaction
for bicycle and train commuters (e.g., Whalen et al., 2013; Handy and
Thigpen, 2019). In a similar way, there are multiple reports that active
travel also tends to yield higher levels of satisfaction (Gatersleben
and Uzzell, 2007; Handy and Thigpen, 2019; Paez and Whalen, 2010; Smith,
2017; St-Louis et al., 2014; Whalen et al., 2013), In contrast, public
transport users often assess their experience more negatively (Abenoza
et al., 2017; De Vos et al., 2016; Gatersleben and Uzzell, 2007; Handy
and Thigpen, 2019; Paez and Whalen, 2010). Multi-modal trips also
influence satisfaction levels; for instance, individuals tend to report
higher levels of satisfaction with the mode they use -- perhaps as a
form of \emph{post hoc} validation of their choices or circumstance
(Susilo and Cats, 2014).

@@

The literature on SWB has demonstrated a relationship between people's
perceptions and satisfaction with their daily travel (e.g., Smith, 2017;
Mokhtarian and Pendyala, 2018; St-Louis et al., 2014). Scholars have
shown that accessibility is among the most studied factors that
influences subjective wellbeing (Delbosc, 2012), and activities have a
direct impact on travel satisfaction (Jakobsson Bergstad, Gamble, T.
Gärling, et al., 2011). Delbosc (2012, p. 28), for instance, has
summarized the most significant influences on psychological wellbeing:
poverty and employment, meaningful relationships and health. However,
understanding the components that affect these perceptions implies the
differentiation between affective (also named as symbolic-affective) and
instrumental values (Jakobsson Bergstad, Gamble, Hagman, et al., 2011).
Steg et al.~(2011) have compared symbolic-affective and
instrumental-reasoned motives for car-use, and similar findings are
reported in other studies (e.g., Gatersleben and Uzzell, 2007; Lois and
Lopez-Saez, 2009). Previous research also demonstrates how
socio-demographic factors affect the levels of SWB, including income
(Clark and Oswald, 1996; Ferrer-i-Carbonell, 2005), education and
unemployment (Argyle et al., 1999), age (Diener and Eunkook Suh, 1997),
and gender (Tesch-Römer et al., 2008). However, further research beyond
the determinants of trip satisfaction is needed to understand how these
socio-demographic variables connect with the affective reactions to
various modes of travel (St-Louis et al., 2014).

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v3_files/figure-latex/plot-descriptive-statistics-1.pdf}
\caption{\label{fig:descriptive-statistics}Descriptive statistics of the
sample}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v3_files/figure-latex/figure-primary-mode-travel-1.pdf}
\caption{\label{fig:primary-mode-travel}Frequency of primary mode used
for regular commute; top panel: all modes, bottom panel: aggregated
modes}
\end{figure}

@@

As seen in the table, there are five individual attributes that we
analyze. Three of these are socio-economic and demographic, namely age,
level of education, and income. The other two are transportation
related, namely primary mode of travel and commute time. The frequency
tables are tested in every case using \(\chi^2\) tests of independence
(\(p\)-values are reported in the table; lower \(p\)-values mean that
the null hypothesis of independence can be rejected with greater
confidence). It is interesting to note that the only category for which
all affective values are significant (\(p\leq0.05\)) is mode of travel.

@@

Turning now to typical commute time, four affective values show
significant differences with \(p\leq0.10\): Freedom, Happiness, Luxury,
and Status. Perhaps not surprisingly, dissonance is more frequent among
people whose typical commutes are longer. This is in line with previous
findings: both St-Louis et al.~(2014) and Smith (2017) report that
commute satisfaction tends to decline with longer commutes, and Handy
and Thigpen (2019) found that commute distance was a negative covariate
of commute satisfaction.

@@

\hspace{1em}> 54 & 74.00 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 72.92 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 68.75 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 76.19 & \multirow{-3}{*}{\centering\arraybackslash 0.4095} & 78.57 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 66.67 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Education}}\\
\hspace{1em}K-12 & 85.98 &  & 92.16 &  & 89.22 &  & 79.27 &  & 95.10 &  & 92.39 & \\

@@

\hspace{1em}Graduate & 78.57 & \multirow{-3}{*}{\centering\arraybackslash 0.9063} & 85.19 & \multirow{-3}{*}{\centering\arraybackslash 0.7005} & 84.62 & \multirow{-3}{*}{\centering\arraybackslash 0.9694} & 86.36 & \multirow{-3}{*}{\centering\arraybackslash 0.3608} & 76.00 & \multirow{-3}{*}{\centering\arraybackslash 0.0058} & 79.17 & \multirow{-3}{*}{\centering\arraybackslash 0.013}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Income}}\\
\hspace{1em}Low & 86.51 &  & 82.11 &  & 88.62 &  & 80.19 &  & 88.71 &  & 88.70 & \\

@@

\hspace{1em}High & 79.19 & \multirow{-3}{*}{\centering\arraybackslash 0.5755} & 90.97 & \multirow{-3}{*}{\centering\arraybackslash 0.2264} & 85.82 & \multirow{-3}{*}{\centering\arraybackslash 0.9698} & 93.69 & \multirow{-3}{*}{\centering\arraybackslash 0.0642} & 73.68 & \multirow{-3}{*}{\centering\arraybackslash 0.0204} & 70.31 & \multirow{-3}{*}{\centering\arraybackslash 0.0137}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Mode}}\\
\hspace{1em}Car & 58.93 &  & 78.90 &  & 70.91 &  & 97.96 &  & 41.00 &  & 30.69 & \\

@@

\hspace{1em}Other & 91.30 & \multirow{-4}{*}{\centering\arraybackslash $< 0.001$} & 86.96 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 91.30 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 93.33 & \multirow{-4}{*}{\centering\arraybackslash 0.0045} & 72.73 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 90.00 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Commute Time}}\\
\hspace{1em}< 20 min & 65.93 &  & 82.95 &  & 77.27 &  & 86.11 &  & 73.49 &  & 67.09 & \\

@@

\hspace{1em}> 60 min & 95.93 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 92.17 & \multirow{-4}{*}{\centering\arraybackslash 0.6115} & 93.28 & \multirow{-4}{*}{\centering\arraybackslash 0.0394} & 83.02 & \multirow{-4}{*}{\centering\arraybackslash 0.9225} & 91.15 & \multirow{-4}{*}{\centering\arraybackslash 0.0925} & 90.09 & \multirow{-4}{*}{\centering\arraybackslash 0.0158}\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v3_files/figure-latex/figure-bar-plots-by-attribute-1.pdf}
\caption{\label{fig:bar-plots-by-attribute}Plots for affective values;
in the y-axis are the number of cases by primary mode of transportation,
and in the x-axis are the modes selected for each affective value}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v3_files/figure-latex/figure-bar-plots-by-attribute-and-age-1.pdf}
\caption{\label{fig:bar-plots-by-age}Plots for affective values by age;
in the y-axis is the proportion in the interval {[}0, 1{]}, and the
x-axis is the mode selected for each value (p-values are for
Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v3_files/figure-latex/figure-bar-plots-by-attribute-and-education-1.pdf}
\caption{\label{fig:bar-plots-by-education}Plots for affective values by
level of education; in the y-axis is the proportion in the interval
{[}0, 1{]}, and the x-axis is the mode selected for each value (p-values
are for Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v3_files/figure-latex/figure-bar-plots-by-attribute-and-income-1.pdf}
\caption{\label{fig:bar-plots-by-income}Plots for affective values by
level of income; in the y-axis is the proportion in the interval {[}0,
1{]}, and the x-axis is the mode selected for each value (p-values are
for Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

Among car users, Freedom is associated with the car in all but the
longest commutes (\textgreater60 min), in which case active travel is
more frequently identified as the mode that evokes feelings of Freedom.
The associations of active travel with Enjoyment and Happiness are
consistently high among car users, although car becomes more frequently
associated with this value for longer commutes. As we have seen before,
active travelers seldom experience dissonance with respect to Freedom,
Enjoyment, and Happiness, and if anything the strength of this
association increases with longer commutes. Public transportation users
again are split in their preference towards active travel and car as the
modes more likely to evoke feelings of Freedom, Enjoyment, and
Happiness, and there are only relatively minor differences in these
preferences by length of commute. The last affective value that displays
significant differences in dissonance by length of typical commute is
Status. The car tends to dominate as the mode more frequently associated
with this value among car and public transportation users. Active
travelers, on the other hand, associate status more strongly with short
or very long commutes: only when commutes are shorter than 20 minutes or
longer than 60 minutes do active travelers associate feelings of Status
with the car.

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v3_files/figure-latex/figure-bar-plots-by-attribute-and-travel-time-1.pdf}
\caption{\label{fig:bar-plots-by-travel-time}Plots for affective values
by commute time; in the y-axis is the proportion in the interval {[}0,
1{]}, and the x-axis is the mode selected for each value (p-values are
for Cochran-Mantel-Haenszel Chi-Squared Test)}
\end{figure}

@@

The premise of this study is that a key component of subjective
wellbeing is the affective reaction to experience (see the definition of
wellbeing offered by the OECD; -@ OECD, 2013). In this paper we
investigated mode dissonance from the perspective of six affective
values. The research presented here was based on a sample of travelers
in Santiago, the capital of Chile. Participants in this research were
asked about their typical mode of travel, and then about the mode or
modes that they associate with each of the six affective values.
Analysis using hypothesis testing (tests of independence) and
visualization techniques (bar charts with faceting) uncovered
interesting patterns. Some of our findings are well aligned with
previous research; for example, active travelers experience less
dissonance than car users, and users of public transportation experience
the most dissonance of all. However, by considering affective values
separately instead of aggregating them into a single indicator of
subjective wellbeing, we manage to preserve greater granularity with
respect to various responses than most studies. This is important
because hedonic/eudaimonic values are more frequently related to active
travel (Freedom, Enjoyment, Happiness), while Poverty is more frequently
related to public transportation and active travel. Luxury and Status,
on the other hand, are more frequently associated to car.

@@

@@

\leavevmode\hypertarget{ref-Bergstad2011subjective}{}%
Jakobsson Bergstad, C., Gamble, A., Gärling, T., Hagman, O., Polk, M.,
Ettema, D., Friman, M., Olsson, L.E., 2011. Subjective well-being
related to satisfaction with daily travel. Transportation 38, 1--15.

\leavevmode\hypertarget{ref-Bergstad2011affective}{}%
Jakobsson Bergstad, C., Gamble, A., Hagman, O., Polk, M., Garling, T.,
Olsson, L.E., 2011. Affective-symbolic and instrumental-independence
psychological motives mediating effects of socio-demographic variables
on daily car use. Journal of Transport Geography 19, 33--38.
doi:\href{https://doi.org/10.1016/j.jtrangeo.2009.11.006}{10.1016/j.jtrangeo.2009.11.006}
